Galois Slicing as Differentiation

### 2025-07-11 main/approx-as-tangents.tex
12.2.8 -> 12.5.9 of Amadio and Curien.
Fix Amadio and Curien theorem refs.
[ tex ] little tweaks: pretty sure it is Thm 12.5.10

\begin{proof}
  \AGDA. See \citet[Theorem 12.2.8]{amadio-curien} for the
  case when the posets are also cpos. The crucial technical step,
  identified by \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq_{\mathrm{ext}} g$ iff
  $\forall x. f(x) \sqsubseteq g(x)$) but instead the stable ordering:
  $f \sqsubseteq_{\mathrm{st}} g$ iff $f \sqsubseteq_{\mathrm{ext}} g$ and
  $\forall x, x'.\,x \sqsubseteq x' \Rightarrow f(x) = f(x') \land
  g(x)$.
\end{proof}

@@

The category of L-posets and stable functions is not cartesian
closed. To make it so, we would need to require that the principal
downsets $\downset{x}$ are \emph{complete}
lattices. \citet[Theorem 12.FIXME]{amadio-curien} details the
proof. Intuitively, to generate the best approximation of an input
value for a function, we need to take the infimum over all possible
input values.

@@

\begin{enumerate}
\item Even without completeness, in Bounded meet posets and L-posets,
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. We discuss this further in
  \secref{conclusion}.) In \exref{intervals-and-maxima-elements},
  maximal elements could be taken to be the ``proper values''. One
  idea is to restrict to conditionally multiplicative or stable
  functions that preserve maximal elements. However, this idea fails
  at higher order: functions that take maximal elements to maximal
  elements are not themselves maximal elements. We could devise a
  category of L-posets with totality predicates (which would pick out
  maximal elements at first-order) and totality preserving functions,
  but we prefer a more direct method of separating values proper from
  their approximations using the \emph{Category of Families}
  construction we detail in \secref{models-of-total-gps:fam}.
\item A more practical reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infinma of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} shows
  how to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}

>>>>

\begin{proof}
  \AGDA. See \citet[Theorem 12.1.9]{amadio-curien} for the
  case when the posets are also cpos. The crucial technical step,
  identified by \citet{berry79}, is that the ordering on conditionally
  multiplicative functions is not the extensional ordering
  ($f \sqsubseteq_{\mathrm{ext}} g$ iff
  $\forall x. f(x) \sqsubseteq g(x)$) but instead the stable ordering:
  $f \sqsubseteq_{\mathrm{st}} g$ iff $f \sqsubseteq_{\mathrm{ext}} g$ and
  $\forall x, x'.\,x \sqsubseteq x' \Rightarrow f(x) = f(x') \land
  g(x)$.
\end{proof}

@@

The category of L-posets and stable functions is not cartesian
closed. To make it so, we would need to require that the principal
downsets $\downset{x}$ are \emph{complete} lattices. \citet[Theorem
12.5.10]{amadio-curien} details the proof. Intuitively, to generate
the best approximation of an input value for a function, we need to
take the infimum over all possible input values.

@@

\begin{enumerate}
\item Even without completeness, in bounded meet posets and L-posets
  values and their approximations live in the same set. However, in
  the \emph{total} query language we wish to model in
  \secref{language}, we are not directly interested in the behaviour
  of programs on approximations as we would be for partial programs
  with general recursion. (Moreover, in the light of
  \exref{intervals-and-maxima-elements} it is not clear whether
  approximations for partiality and approximations for stability ought
  to be the same thing. We discuss this further in
  \secref{conclusion}.) In \exref{intervals-and-maxima-elements},
  maximal elements could be taken to be the ``proper values''. One
  idea is to restrict to conditionally multiplicative or stable
  functions that preserve maximal elements. However, this idea fails
  at higher order: functions that take maximal elements to maximal
  elements are not themselves maximal elements. We could devise a
  category of L-posets with totality predicates (which would pick out
  maximal elements at first-order) and totality preserving functions,
  but we prefer a more direct method of separating values proper from
  their approximations using the \emph{Category of Families}
  construction as we explain in \secref{models-of-total-gps:fam}.
\item A more practical reason is that we wish to formalise our
  construction in the proof assistant Agda \cite{agda} in order to get
  an \emph{executable} model of the language in
  \secref{language}. Agda's type theory is both predicative and
  constructive. Predicativity means that it does not have complete
  lattices in the classical sense: for a type $X$ we can only get
  suprema and infinma of families in a lower universe level than
  $X$. This is not necessarily a problem, as \citet{dejong21} show how
  to develop a large amount of domain theory in a predicative
  setting. However, due to constructivity the lifting construction
  (due to \citet{escardo-knapp}) requires a proof of definedness to
  extract a value. Since we are modelling a total language, we prefer
  to have a model that computes directly.
\end{enumerate}
